Higher-Order unification for free!

### 2024-04-10 paper/main.tex
section 1 and 2

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitated this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone in proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
In particular we want to leverage the Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances'', hence
we can expect that Elpi is a good fit for
implementing such as form of automation. In this paper we focus on one aspect of
this work, namely \emph{how to reuse the higher order unification procedure
of the meta language in order to implement a type-class solver for the object
language}. As it turns out, re-using the unification of the meta language is
not a trivial task.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three designated \coqIn{Instances} state that:
1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n}, is finite;
2) the predicate \coqIn{nfact n nf}, linking a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with rules
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular by solving the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and hence admits a unique
solution \coqIn{P = ~$\lambda$~x.nfact x 3}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of rules. Elpi comes with
an Higher Order Abstract Syntax~\cite{10.1145/53990.54010} datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets denote a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

\begin{elpicode}
all (con"t") y\ app[con"nfact", y, con"3"]
\end{elpicode}

@@

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal g given below

@@

\noindent
In this paper we study a more sophisticated encoding of rules that, on a first
approximation, would shape \ref{clause:r3} as follows:

@@

\noindent
Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
back to the domain \elpiIn{tm} of Coq terms:

\begin{elpicode}
P = lam A a\ app [con"nfact", a, con"3"]
\end{elpicode}

@@

\begin{elpicode}
decision (app[                    P                   , w])
decision (app[ lam A (a\ app [con"nfact", a, con"3"]) , w])
\end{elpicode}

@@

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories and
unification procedures of the meta language and the object language.

The equational theory of CIC is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi is strikingly similar,
since it it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves problems in
\llambda as well.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different.
% For example:\\
% %
% \vspace{4pt}
% {
% \setlength{\tabcolsep}{2pt}
% \begin{tabular}{lcl|}
% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
% \end{tabular}
% \phantom{mn}
% \begin{tabular}{lcl}
% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
% \end{tabular}
% }
% \vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax as
per \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure} %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

The first four rules are common to both equalities
and just define the usual congruence over terms. Since
we use an HOAS encoding they also capture $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

We write $\sigma = \{~ X \mapsto t ~\}$ the substitution that assigns
the term $t$ to the variable $X$.
We write $\sigma t$ for the application of
the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
$X$ is a set of terms.
We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$.
The domain of a substitution is the set of unification variables for which
it provides an assignment.
We write $\sigma \cup \sigma'$ set union to denote the concatenation of
two substitutions whose domains are disjoint.
We shall use $\rho$ for \Fo{} substitutions,
and $\sigma$ for the \Ho ones.
\todo{can be made a little more formal}
For brevity, in this section we consider
the substitution for \Fo{} and \Ho{} identical.
We defer to \cref{sec:grounwork} a more precise description
pointing out theirs differences.

@@

Although we provide an implementation of \Ue in the supplementary material
(that we used for testing purposes) we only describe its signature here.

@@

%
% To state precisely the problem we solve we need a \Fo{} representation
% of CIC terms and a \Ho one. We extend the equality over ground
% terms to open ones with the reflexivity of unification variables (that is
% unification variable X is equal to itself, not to another one such a Y).
% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
% We call \Uo the unification procedure we want to implement and
% \Ue the one provided by the meta language.

Although we provide an implementation of \Uo{} in \cref{sec:founif},
our real goal is the simulation of an entire logic program.

\subsection{The problem: Logic Program Simulation}

@@

\noindent 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each term $s \in \Fo{}$ into a term $t \in \Ho{}$. We write this
step $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:compilation,sec:eta,sec:beta}, here we just point
out that it additionally a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

%\cref{sec:beta}
%XXX permuting hrun does not change the final result if check does not fail eagerly
% XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to a non MGU too early

% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
% %$\foUnifPb = \{ s_1, s_2 \}$ 
% as follows:
% %
% $$
% \begin{array}{l}
% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
% \end{array}
% $$

% \begin{proposition}[Properties of \Uo{} in \llambda]
% \begin{gather}
%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
% \end{gather}
% \end{proposition}

%In addition to correctness and completeness in \llambda,
We also claim that \hrun
handles terms outside \llambda in the following sense:

\begin{proposition}[Fidelity recovery]
\begin{gather}
  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
\end{gather}
\end{proposition}

\noindent
Property \ref{prop:compilation-nllam} states that two terms for which
there is a unifier (given by an oracle and not necessarily a most general one),
then the compiler generates two terms that unify in \Ho.\\
%In other words the compiler moves out of the way all problematic terms.
Property \ref{prop:step-nllam} says that if the two terms involved in a step
re-enter \llambda, then \hstep succeeds. This is a typical example
in which the order of the unification problems in a logic-program run
does matter. The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda. In other words
\cref{prop:fidelity} \todo{...}

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms. We now define ``problematic'' and ``well behaved''
more formally.

@@

\noindent
An example of term $t$ in \maybebeta{} is the application $F \appsep x$.
This term is problematic since the application node of
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$ the term
head constructor may become a $\lambda$, or a constant or stay an application.

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $t$ such that $\exists \rho, \rho t$
  is an eta expansion.
\end{definition}

\noindent
An example of term $t$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
makes $\rho t = \lambda x.\lambda y.f \appsep x \appsep y$
that is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant or an application.

@@

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but cannot be handled directly by the unification of the meta language, that
is only required to handle terms in \llambda.

@@

\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\notllambda{} ~\cup~ \maybeeta{} ~\cup~ \maybebeta{})
$$
\end{definition}

\noindent

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\hoUnifPb) \land
  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
\wellb(\sigma\hoUnifPb) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \hoUnifPb)
\end{array}
$$
\end{proposition}

\noindent
Proposition \ref{prop:nf} is key to prove \cref{prop:simulation,prop:fidelity}:
informally it says that the problematic terms moved on the side by the compiler
are not put back by \hstep, hence \Ue{} can operate properly.
In \cref{sec:compilation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and granting \cref{prop:simulation,prop:fidelity}.

% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) problematic terms.

@@

% \noteD[inline]{Below the proof of \cref{prop:nf}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa $l$, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:nf}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:nf}
  \label{lemma:unif-wellb}
\end{lemma}

>>>>

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitates this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone of proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
In this paper we focus on one aspect of
this work, precisely \emph{how to reuse the higher order unification procedure
of the meta language in order to simulate a higher order logic program
for the object language}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three designated type-class instances state that:
1) the type of natural numbers smaller than \coqIn{n}, called \coqIn{fin n},
is finite;
2) the predicate \coqIn{nfact n nf}, relating a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and admits a unique
solution $\customlabel{solution:intro}{\sigma}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of instances as rules (a.k.a. clauses).
Elpi comes equipped with
an Higher Order Abstract Syntax (HOAS~\cite{10.1145/53990.54010}) datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets delimit a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

\begin{elpicode}
all (con"t") y\ app [con"nfact", y, con"3"]
\end{elpicode}

@@

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

@@

\noindent
In this paper we study a more sophisticated encoding of CIC terms and rules
that, on a first approximation, would reshape \ref{clause:r3} as follows:

@@

\noindent
Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
back to the domain \elpiIn{tm} of Coq terms, obtaining the expected solution
\ref{solution:intro}:

\begin{elpicode}
P = lam A x\ app [con"nfact", x, con"3"]
\end{elpicode}

@@

\begin{elpicode}
decision (app [                   P                  , w])
decision (app [lam A (a\ app [con"nfact", a, con"3"]), w])
\end{elpicode}

@@

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, that in turns makes the 
unification procedures of the meta language weak.\\
The equational theory of the meta language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher order
problems in the pattern fragment. Although 
the equational theory of CIC is much richer, for efficiency and predictability
reasons automatic proof search procedure typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs some care in order to
simulate a logic program in CIC using the unification of Elpi.

% For example:\\
% %
% \vspace{4pt}
% {
% \setlength{\tabcolsep}{2pt}
% \begin{tabular}{lcl|}
% \elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
% \elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
% \elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
% \end{tabular}
% \phantom{mn}
% \begin{tabular}{lcl}
% \elpiIn{P x} & \Ue{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
% \elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
% \end{tabular}
% }
% \vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax as
per \cref{code:common-terms}.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

The first four rules are common to both equalities
and define the usual congruence over terms. Since
we use an HOAS encoding they also capture $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

We write $\sigma = \{~ X \mapsto t ~\}$ for the substitution that assigns
the term $t$ to the variable $X$.
We write $\sigma t$ for the application of
the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
$X$ is a set of terms.
We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$.
The domain of a substitution is the set of unification variables for which
it provides an assignment.
We write $\sigma \cup \sigma'$ to denote the concatenation of
two substitutions whose domains are disjoint.
We shall use $\rho$ for \Fo{} substitutions,
and $\sigma$ for the \Ho ones.
For brevity, in this section we consider
the substitution for \Fo{} and \Ho{} identical.
We defer to \cref{sec:grounwork} a more precise description
pointing out theirs differences.

@@

Although we provide an implementation of the meta-language unification \Ue
in the supplementary material (that we used for testing purposes) we only
describe its signature here.

@@

Even if we provide an implementation of the object-language unification
\Uo{} in \cref{sec:founif}, our real goal is the simulation of an entire
logic program.

\subsection{The problem: logic-program Simulation}

@@

\noindent
\todo{say that backtracking is not important} 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each \Fo{} term in \foUnifPb into a \Ho{} term $t$.
We write this step $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:simulation,sec:eta,sec:beta}, here we just point
out that it additionally produce a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho to variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

@@

We also claim that \hrun handles terms outside \llambda in the following sense:

\begin{proposition}[Fidelity recovery]\label{prop:fidelity-recovery} 
In the context of \hrun, if $\sigma_{p-1} \hoUnifPb_{p} \in \llambda$ then
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p} \Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p}
$$
\end{proposition}

In other words if the two terms involved in a step
re-enter \llambda, then \hstep and \fstep are again related, even if
$\foUnifPb \not\subseteq \llambda$ and hence \cref{prop:fidelity} does not apply.

This property has a practical relevance since in many logic programming
implementations, including Elpi, the order in which unification problems
are tackled does matter.
The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda.

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled to a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
with an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms.\\
We now define ``problematic'' and ``well behaved'' more formally.

@@

\noindent
An example of term \maybebeta{} is the application $F \appsep x$.
This term is problematic since the application node of
its syntax tree cannot be used to justify a
unification failure, i.e. by properly instantiating $F$ the term
head constructor may become a $\lambda$, or a constant or stay an application.

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $s$ such that $\exists \rho, \rho s$
  is an eta expansion.
\end{definition}

\noindent
An example of term $s$ in \maybeeta{} is
$\lambda x.\lambda y.F \appsep y \appsep x$
since the substitution
$\rho = \{ F \mapsto \lambda a.\lambda b.f \appsep b \appsep a\}$
makes $\rho s = \lambda x.\lambda y.f \appsep x \appsep y$
that is the eta long form of $f$. This term is problematic since
its leading $\lambda$ abstraction cannot justify a
unification failure against a constant $f$.

@@

\noindent
These terms are problematic for the very same reason terms in \maybebeta are,
but cannot be handled directly by the unification of the meta language, that
is only required to handle terms in \llambda. Still, there exists a
substitution $\rho$ such that $\rho s \in \llambda$.

@@

\newcommand{\wellb}{\ensuremath{\mathcal{W}}\xspace}
\begin{definition}[Well behaved set]
Given a set of terms $X \subseteq \Ho{}$,
$$
\wellb(X) \Leftrightarrow \forall t \in \subterm{X},
t \not\in (\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{})
$$
\end{definition}

\noindent
We claim our compiler validates the following property:

\begin{proposition}[\wellb-enforcing]\ref{prop:w-enforcing}
  Given two terms $s_1$ and $s_2$, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-w}
  $$
\end{proposition}

\noindent
Note that the property holds for any substitution, it could be given by an
oracle and not necessarily a most general one, and for any terms,
in particular terms in $\maybebeta{} ~\cup~ \maybeeta{} ~\cup~ \notllambda{}$.

\begin{proposition}[\wellb{}-preservation]\label{prop:w-preservation}
$\forall \hoUnifPb, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\begin{array}{l}
\wellb(\sigma\hoUnifPb) \land
  \sigma\hoUnifPb_{p_l} \Ue \sigma\hoUnifPb_{p_r} \mapsto {\sigma'}
  \Rightarrow \wellb(\sigma' \hoUnifPb)\\
\wellb(\sigma\hoUnifPb) \land
  \progress(\linkStore,\sigma) \mapsto (\_,\sigma')
  \Rightarrow \wellb(\sigma' \hoUnifPb)
\end{array}
$$
\end{proposition}

\noindent
Proposition \ref{prop:w-preservation} is key to prove \cref{prop:simulation,prop:fidelity}:
informally it says that the problematic terms moved on the side by the compiler
are not put back by \hstep, hence \Ue{} can continue to operate properly.
In \cref{sec:simulation,sec:eta,sec:beta}
we describe how the compiler recognizes terms in \maybebeta, \maybeeta and
\notllambda and how \progress takes care of them preserving \wellb
and granting \cref{prop:simulation,prop:fidelity,prop:fidelity-recovery}.

% Note that proposition \ref{prop:w-preservation} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) problematic terms.

@@

% \noteD[inline]{Below the proof of \cref{prop:w-preservation}, ho usato 3 lemmi ausiliari,
% forse si può compattare in una prova più piccola?}
\begin{lemma}
  Given a \linketa $l$, the unification done by \progressetaleft is between
  terms in $\wellb$  
  \label{lemma:unif-eta-1}
\end{lemma}

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, which is  $\wellb(\sigma)$ (by
  \cref{prop:w-preservation}). $\lhs \in \sigma$, therefore $\wellb(\lhs)$. By
  \progressetaleft, if 1) \lhs is a name, a constant or an
  application, then, $\lambda x.\lhs \appsep x$ is unified with \rhs. By
  \cref{inv:link-eta-right} and \cref{lemma:unif-eta-aux}, $\rhs = \lambda x. t$ and $\wellb(t)$.
  Otherwise, 2) \lhs has \elpiIn{lam} as functor. In both cases, unification is
  performed between terms in \wellb.
\end{proof}

@@

\begin{lemma}
  The introduction of \linketa guarantees \fullref{prop:w-preservation}
  \label{lemma:unif-wellb}
\end{lemma}
